### 2020-04-22 final_report/data.tex
add some references
intermediate replacement of the AE plots, some minor edits

Autoencoders are a family of approaches to feature extraction driven by neural
networks. Faced with many possible alternatives, we implemented a conventional
autoencoder, which relies on a three-layer network that is trained to replicate
the identity mapping. While it follows that the input and output layers of such
network are sized to accommodate the analysed dataset, the
hidden layer---also called the bottleneck---allows for variable number of
neurons that represent a smaller subspace. By scanning
over a range of bottleneck widths and investigating relative changes in the
validation loss, we assess the potential for dimensional reduction.

This approach was applied to two sets of samples: (i) a subset of data obtained
from run~1 and (ii) a subset of a single slice obtained from run~2. Our
expectation was that while the former case would provide meaningful insight into
correlations within the feature space, the latter would validate our
autoencoder implementation by analysing a set of points that are trivially
reducible due to a number of fixed features.

\begin{wrapfigure}{l}{0.4\textwidth}
	\centering
	\vspace{-1ex}
	% TODO: placeholders only, regenerate when time is available
	\begin{subfigure}[b]{\linewidth}
		\centering
		\includegraphics[width=\linewidth]{ae_all_slices}
	\end{subfigure}
	\begin{subfigure}[b]{\linewidth}
		\centering
		\includegraphics[width=\linewidth]{ae_single_slice}
	\end{subfigure}
	\caption{Autoencoder loss in two considered cases of dimensional reduction.}
	\label{fig:autoencoder-loss}
	\vspace{-6ex}
\end{wrapfigure}

The results of both experiments are shown in~\cref{fig:autoencoder-loss}.
Consistent with our motivation, in each plot we can clearly identify a constant
plateau in the region of large dimensionality followed by a point, from which a
steep increase in loss is observed towards lower dimension counts. We consider
this critical point to correspond with the largest viable dimensional reduction
without significant information loss.
With this methodology we find that the autoencoder was able to reduce the
datasets into a subspace of 17~dimensions in the first case and 10~dimensions in
the second case.

Confirming our expectation that in the latter, trivially reducible case the
autoencoder should achieve stricter dimensional reduction, we are
encouraged to believe that our implementation is indeed efficient. However, we
must also conclude that in both examined cases this method failed to produce a
reduction that would be dominate a naïve approach, in which we eliminate
7~dimensions due to overdetermined one-hot-encoded features, 2~dimensions due to
sum-to-one constraints, and in the latter case 7~additional dimensions due to
artificially fixed features. This supports the findings from PCA as well as the
presented semivariograms.

>>>>

Autoencoders~\cite{SCHMIDHUBER201585} are a family of approaches to feature extraction driven by neural
networks. Faced with many possible alternatives, we implemented a conventional
autoencoder, which relies on a three-layer network that is trained to replicate
the identity mapping. While it follows that the input and output layers of such
network are sized to accommodate the analysed dataset, the
hidden layer---also called the bottleneck---allows for variable number of
neurons that represent a smaller subspace. By scanning
over a range of bottleneck widths and investigating relative changes in the
validation loss, we assess the potential for dimensional reduction.

This approach was applied to two sets\footnote{Each set contained
\num{100000}~samples from batches 0-99 of the corresponding runs.} of samples: (i) a subset of data obtained
from run~1 and (ii) a subset of a single slice obtained from run~2. Our
expectation was that while the former case would provide meaningful insights into
correlations within the feature space, the latter would validate our
autoencoder implementation by analysing a set of points that are trivially
reducible due to a number of fixed discrete features.

\begin{wrapfigure}{l}{0.4\textwidth} % TODO: regenerate both plots as described
	\centering
	\vspace{-1ex}
	\begin{subfigure}[b]{\linewidth}
		\includegraphics[width=\linewidth,trim={0 48px 0 0},clip]{ae_mixed}
	\end{subfigure}

	\vspace{-0.2ex}

	\begin{subfigure}[b]{\linewidth}
		\includegraphics[width=\linewidth]{ae_single_slice}
	\end{subfigure}

	\caption{Autoencoder loss scan on the full feature space (top) and a single slice
		(bottom). Dimensional reduction is indicated by a green arrow.}
	\label{fig:autoencoder-loss}
	\vspace{-6ex}
\end{wrapfigure}

The results of both experiments are shown in~\cref{fig:autoencoder-loss}.
Consistent with our motivation, in each plot we can clearly identify a constant
plateau of low error in the region of large dimensionality followed by a point,
from which a steep increase is observed.
We consider this \textit{critical point} to mark the largest viable
dimensional reduction without significant information loss.
With this approach we find that the autoencoder was able to reduce the
datasets into a subspace of 18~dimensions in the first case and 10~dimensions in
the second case.

Confirming our expectation that in the latter, trivially reducible case the
autoencoder should achieve greater dimensional reduction, we are inclined to
believe that our implementation is indeed efficient. However, we
must also conclude that in both examined cases this method failed to produce a
reduction that would be superior to a naïve approach.\footnote{In both tested cases we
	can trivially eliminate 7~dimensions due to overdetermined one-hot-encoded
	features and 2~dimensions due to sum-to-one constraints. Furthermore, in the
	single slice case we may omit 7~additional dimensions due to artificially fixed
features.} This supports the findings from PCA as well as the
presented semivariograms.
